# Snowfall prediction with machine learning methods

The project aims to apply selected machine learning methods to the classification problem. <br>
After analyzing meteorological data collected for Cracow, Poland, a classification will be conducted to determine whether snowfall is expected on a given day in November. The project includes machine learning techniques like Logistic Regression, Decision Tree, and Support Vector Machines, with their outcomes evaluated against each other. Additionally, the interpretability of Logistic Regression will be scrutinized.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(ingredients)
library(glmnet)
library(DALEX)
library(gridExtra)
library(ggplot2)
library(corrplot)
library(pROC)
library(ipred)
library(rpart)
library(randomForest)
library(readr)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(caret)
library(ipred)
library('e1071') 
library(glmnet)
library(dplyr)
library(gbm)
library(ROSE)
library(rmarkdown)
library(kableExtra)
```

```{r, include=FALSE}
data_snow <- read.csv("Krakow_11_snieg_en.csv")
```

### Data Exploration 

The data used in the task was collected from a public dataset provided by IMGW (Polish Institute of Meteorology and Water Management). The observations come from the synoptic station number 350190566, located at Cracow Balice Airport, which is the main station gathering information about weather conditions in Cracow. Measurements are taken daily and collected for one month - November, over the years 2010 - 2023, in total 420 days.

The dataset provided by IMGW includes all parameters of atmospheric conditions recorded at meteorological stations. The data has been cleaned of unnecessary variables, such as: the duration of sleet or the occurrence of lightning, as these phenomena, even over several years, are recorded too rarely to have a significant impact on the study.
Variables that are not considered fundamental atmospheric phenomena have been also removed, such as minimum ground temperature or ground state, as their influence on the weather forecasting model will not be practical.

```{r, echo=FALSE}
f <- function() {
  paged_table(data_snow)
}

f()
```

Variables description:

- `cloudiness`:  degree of  sky covered by clouds on a scale of 0-8, where 0 indicates  no clouds and 8 indicates sky completely covered by clouds

- `wind`: average daily wind speed in m/s

- `temperature_min`: maximum daily air temperature in Celsius degrees 

- `temperature_max`: minimum daily air temperature in Celsius degrees 

- `temperature`: average daily air temperature in Celsius degrees 

- `humidity`: average air water vapor content in %

- `pressure`: average daily pressure in hPa

- `dew`: time of dew occurrence within a day in hours

- `frost`: time of frost occurrence within a day in hours

- `fog`: time of fog occurrence within a day in hours, where fog is a suspension of small water droplets in the lower layer of air reducing visibility to 1 km

- `mist`: time of mist occurrence within a day in hours, where mist is a suspension of small water droplets in the lower layer of air reducing visibility to 1 - 10 km

- `haze`: time of haze occurrence within a day in hours, where haze is a suspension of dust and atmospheric aerosols limiting visibility

- `wind10`: time of wind speed exceeding 10 m/s within a day in hours

- `snow`: amount of daily snowfall in mm

- `snow_cover`: height of snow cover in cm

- `rain`: amount of daily rainfall in mm

```{r, include=FALSE}
data_snow$dew_01 <- ifelse(as.numeric(data_snow$dew) >= 0.5, 1, 0)
data_snow$frost_01 <- ifelse(as.numeric(data_snow$frost) >= 0.5, 1, 0)
data_snow$fog_01 <- ifelse(as.numeric(data_snow$fog) >= 0.5, 1, 0)
data_snow$wind10_01 <- ifelse(as.numeric(data_snow$wind10) >= 0.5, 1, 0)
data_snow$snow_cover_01 <- ifelse(as.numeric(data_snow$snow_cover) >= 0.1, 1, 0)
data_snow$rain_01 <- ifelse(as.numeric(data_snow$rain) >= 0.5, 1, 0)
data_snow$haze_01 <- ifelse(as.numeric(data_snow$haze) >= 0.5, 1, 0)
data_snow$mist_01 <- ifelse(as.numeric(data_snow$mist) >= 0.5, 1, 0)
data_snow$snow_prediction <- ifelse(as.numeric(data_snow$snow) >= 0.1, 1, 0)
```

For data analysis and extraction of endogenous variables, binary values were assigned to the following variables: `dew`, `frost`, `fog`, `wind10`, `snow_cover`, `rain`, `haze`, and `mist`.  A value of 1 is assigned to these variables if their duration exceeds 30 minutes during the day, except for snow cover, which takes a value of 1 after reaching 10 mm of snow cover during the day. These binary values are named accordingly: `dew_01`, `frost_01`, `fog_01`, `wind10_01`, `snow_cover_01`, `rain_01`, `haze_01`, and `mist_01`.

The exogenous variable is `snow_prediction`, which takes a value of 1 in case of snow occurrence more than 0.1 hour during the day.

```{r, echo=FALSE}
createDataSnow <- function() {

  data_snow$cloudiness <- as.numeric(data_snow$cloudiness)
  data_snow$wind <- as.numeric(data_snow$wind)
  data_snow$temperature <- as.numeric(data_snow$temperature)
  data_snow$temperature_min <- as.numeric(data_snow$temperature_min)
  data_snow$temperature_max <- as.numeric(data_snow$temperature_max)
  data_snow$humidity <- as.numeric(data_snow$humidity)
  data_snow$pressure <- as.numeric(data_snow$pressure)
  
  data_snow$dew_01 <- factor(as.factor(data_snow$dew_01), levels=c(0,1), labels=c('Deficit','Dew'))
  data_snow$frost_01 <- factor(as.factor(data_snow$frost_01), levels=c(0,1), labels=c('Deficit','Frost'))
  data_snow$fog_01 <- factor(as.factor(data_snow$fog_01), levels=c(0,1), labels=c('Deficit','Fog'))
  data_snow$wind10_01 <- factor(as.factor(data_snow$wind10_01), levels=c(0,1), labels=c('Deficit','Wind_above_10m/s'))
  data_snow$snow_cover_01 <- factor(as.factor(data_snow$snow_cover_01), levels=c(0,1), labels=c('Deficit','Snow_cover'))
  data_snow$rain_01 <- factor(as.factor(data_snow$rain_01), levels=c(0,1), labels=c('Deficit','Rain'))
  data_snow$haze_01 <- factor(as.factor(data_snow$haze_01), levels=c(0,1), labels=c('Deficit','Haze'))
  data_snow$mist_01 <- factor(as.factor(data_snow$mist_01), levels=c(0,1), labels=c('Deficit','Mist'))
  data_snow$snow_prediction <- factor(as.factor(data_snow$snow_prediction), levels=c(0,1), labels=c('Deficit','Snow'))

  return(data_snow)
}

data_snow <- createDataSnow()

```

```{r, include=FALSE}
data_snow<- subset(data_snow, select = -c(id, city, year, month, day, dew, frost, fog, mist, haze, mist, wind10, snow, snow_cover, rain))
colnames(data_snow)
```

```{r, echo=FALSE}
data_snow %>%
  keep(is.numeric) %>%
  summary()
```
```{r, echo=FALSE}
data_snow %>%
  keep(is.factor) %>%
  summary()
```

```{r, include=FALSE}
data_snow <-data_snow %>%
  mutate(temperature_min = ifelse(is.na(temperature_min), mean(temperature_min, na.rm = TRUE), temperature_min))
```
Missing data values were replaced with their mean.
```{r, echo=FALSE}
data_snow %>%
  keep(is.numeric) %>%
  summary()
```
The exogenous variable is indicating the occurrence of snowfall. Snow is characteristic only for specific atmospheric conditions, which may facilitate finding relationships and the accuracy of models. The following charts illustrate the ratio of days with the occurrence of selected atmospheric phenomena to days without their occurrence in the analyzed dataset. Snow does not occur as often as fog or rain, but it is a more common phenomenon than haze or wind blowing faster than 10m/s.

```{r, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
categorical_vars <- c("snow_prediction", "mist_01", "rain_01", "wind10_01", "haze_01")

gp_categorical <- lapply(categorical_vars, function(x) { 
  ggplot(data = data_snow, aes_string(x = x)) + 
    geom_bar() + 
    labs(title = paste(x, "distribution"), x = x, y = 'Number of days')
})

num_plots <- length(gp_categorical)
num_cols <- 2
num_rows <- ceiling(num_plots / num_cols)

chunks_categorical <- split(gp_categorical, rep(1:num_rows, each = num_cols, length.out = num_plots))

grob_plots_categorical <- lapply(chunks_categorical, function(chunk) {
  marrangeGrob(grobs = lapply(chunk, ggplotGrob), nrow = 1, ncol = num_cols)
})

invisible(print(grob_plots_categorical))
```

The analysis of binary variables and their relationship to the dependent variable was run using the Chi-square independence test. Setting hypotheses for the variable `rain_01`:

*H0: Snowfall is not dependent on the occurrence of rainfall.*

*H1: Snowfall is dependent on the occurrence of rainfall.*

Similarly, the same procedure was applied to the other categorical variables. The tests revealed that variables:`wind10_01`, `haze_01` and `mist_01` are not statistically significant. This might be due to the very rare occurrence of wind above 10m/s and haze over the years. On the other hand, mist occurs frequently enough that it does not distinctly define the occurrence of snowfall.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
chi_results <- c(
  chisq.test(data_snow$rain_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$dew_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$frost_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$fog_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$wind10_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$snow_cover_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$haze_01, data_snow$snow_prediction)$p.value,
  chisq.test(data_snow$mist_01, data_snow$snow_prediction)$p.value
)
```

```{r, echo=FALSE}
chi_results <- data.frame(chi_results)
chi_results <- t(chi_results)
colnames(chi_results) <- c("rain_01", "dew_01", "frost_01", "fog_01", "wind10_01", "snow_cover_01", "haze_01", "mist_01")
rownames(chi_results) <- c("p-value")

pvalue_results <- head(chi_results)

knitr::kable(pvalue_results, 
             align = "cccccccc",
             caption = "P-values from a Chi-square test on categorical variables",
             digits = 3) %>%
kable_styling(latex_options = "striped", full_width = F) %>%
column_spec(1, width = "8cm") %>% 
column_spec(2, width = "6cm") %>%
column_spec(3, width = "6cm") %>%
column_spec(4, width = "6cm") %>%
column_spec(5, width = "6cm") %>%
column_spec(6, width = "6cm") %>%
column_spec(7, width = "6cm") %>%
column_spec(8, width = "6cm")
```

The following graph depicts a visualization of the correlation relationships between variables. Based on the graph, it is observable that humidity is positively correlated with cloudiness, wind is positively correlated with cloudiness.

On the other hand, maximum temperature is negatively correlated with cloudiness, maximum temperature is negatively correlated with humidity, and humidity is negatively correlated with wind.

```{r, echo=FALSE, message=FALSE, warning=FALSE,results='hide'}
correlation_matrix <- cor(data_snow[,c( "cloudiness", "wind", "temperature", "humidity", "pressure", "temperature_max")])

print(correlation_matrix)

corrplot(correlation_matrix, method = "circle")
```

Visualizing the significance of predictors is possible through box plots and density plots. These plots help to understand the alignment of numerical variable values with the predicted variable of snowfall. 

In the box plots, variables such as pressure and humidity were not significantly different between the distribution with snowfall and without snowfall.
Box plots for weather variables such as temperature, minimum temperature, maximum temperature, wind, and cloudiness show significant differences in relation to the dependent variable. For instance, the box plot for average wind speed shows a higher median and wider range of wind speed values on days with snowfall.

Also, density plots showed that humidity and pressure do not exhibit significant differences on days with snowfall compared to days without snowfall. As a result, these variables will not be taken into account during model creation. To avoid duplicating several variables describing the same atmospheric phenomenon, average daily temperature has been chosen, defined as the variable temperature. It exhibits the greatest variability in the density plot between days with and without snowfall.


```{r, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
numeric_vars <- c("temperature", "temperature_min", "temperature_max", "humidity", "pressure", "cloudiness", "wind")

gp_numeric <- lapply(numeric_vars, function(var) {
  data_snow %>%
    filter(!is.na(!!sym(var))) %>%
    ggplot(aes_string(x = "snow_prediction", y = var)) +
    geom_boxplot(fill = 'red', alpha = 0.7) +
    scale_y_continuous(name = switch(var,
                                      "temperature" = "Average daily temperature",
                                      "temperature_min" = "Average daily minimum temperature",
                                      "temperature_max" = "Average daily maximum temperature",
                                      "humidity" = "Humidity",
                                      "pressure" = "Pressure",
                                      "cloudiness" = "Cloudiness",
                                      "wind" = "Wind"),
                      breaks = switch(var,
                                      "temperature" = seq(-15, 20, 5),
                                      "temperature_min" = seq(-15, 20, 5),
                                      "temperature_max" = seq(-15, 20, 5),
                                      "humidity" = seq(20, 100, 20),
                                      "pressure" = seq(900, 1050, 50),
                                      "cloudiness" = seq(0, 8, 2),
                                      "wind" = seq(0, 15, 5)),
                      limits = switch(var,
                                      "temperature" = c(-15, 20),
                                      "temperature_min" = c(-15, 20),
                                      "temperature_max" = c(-15, 20),
                                      "humidity" = c(20, 100),
                                      "pressure" = c(900, 1050),
                                      "cloudiness" = c(0, 9),
                                      "wind" = c(0, 15))) +
    scale_x_discrete(name = "Is it snowing?") +
    ggtitle("") +
    theme_bw()
})

chunks_numeric <- split(gp_numeric, ceiling(seq_along(gp_numeric) / 2))  # Split plots into chunks of 4

grob_plots_numeric <- lapply(chunks_numeric, function(chunk) {
  marrangeGrob(grobs = lapply(chunk, ggplotGrob), nrow = 1, ncol = 2)
})

invisible(print(grob_plots_numeric))
```
```{r, echo=FALSE, message=FALSE, results='hide'}
numeric_vars <- c("temperature", "temperature_min", "temperature_max", "humidity", "pressure", "cloudiness", "wind")

gp <- lapply(numeric_vars, function(x) { 
  ggplot(data = data_snow, aes(x = !!sym(x), col = snow_prediction)) + 
    geom_density() + 
    xlab(x) + 
    ggtitle(paste(x, "density", sep = " "))
})

chunks <- split(gp, ceiling(seq_along(gp) / 4))  # Split plots into chunks of 4

grob_plots <- lapply(chunks, function(chunk) {
  marrangeGrob(grobs = lapply(chunk, ggplotGrob), nrow = 2, ncol = 2)
})

invisible(print(grob_plots))
```


The variables included in the model after data exploration:
```{r, include=FALSE}
data_snow<- subset(data_snow, select = -c(temperature_max, temperature_min, pressure, humidity, wind10_01, mist_01, haze_01))
colnames(data_snow)
```

```{r, echo=FALSE}
f_2 <- function() {
  paged_table(head(data_snow, 5))
}

f_2()
```

### Training and testing set 

```{r, echo=FALSE}
numbers <- table(data_snow$snow_prediction)
share <- prop.table(numbers) * 100

cat("Number of positive observations (snow):", numbers["Snow"], "\n")
cat("Number of negative observations (deficit):", numbers["Deficit"], "\n")
cat("Percentage of positive observations (snow):", share["Snow"], "%\n") 
cat("Percentage of negative observations (deficit):", share["Deficit"], "%\n")
```
Division into training and testing set in the ratio of 70:30.
```{r, echo=FALSE}
set.seed(1234)
sample_set <- sample(nrow(data_snow), round(nrow(data_snow)*.70), replace = FALSE)
data_train <- data_snow[sample_set, ]
data_test <- data_snow[-sample_set, ]
```

Training set:
```{r, echo=FALSE}
table(data_train$snow_prediction)
```

Testing set:
```{r, echo=FALSE}
table(data_test$snow_prediction)
```
Standardization was performed on the data. 
```{r, include=FALSE}
normalize <- function(x){
  return((x-min(x))/(max(x)-min(x)))
}

data_train <- data_train %>%
  mutate(cloudiness = normalize(cloudiness)) %>%
  mutate(wind = normalize(wind)) %>%
  mutate(temperature = normalize(temperature))
```

```{r, echo=FALSE}
data_train %>%
  keep(is.numeric) %>%
  summary()
```
It has been observed that the share of the positive class is greater than the negative class. In order to balance the data set, oversampling was performed on the training set. The share of observations is the same in the positive and negative class now.

Oversampled training set: 
```{r, echo=FALSE}
oversampled_data_train <- ovun.sample(snow_prediction ~ ., data = data_train, method = "over", N = 2*nrow(subset(data_train, data_train$snow_prediction == "Deficit")))$data
table(oversampled_data_train$snow_prediction)
```
### Logistic regression
One of the regression methods used in statistics when the dependent variable is on a dichotomous scale (takes only two values).

##Logistic regression on train data 

```{r}
logistic_model_evaluation <- function(train_data, test_data, formula, threshold = 0.5) {
  logistic_train <- glm(formula, data = train_data, family = "binomial")
  
  logistic_predict_train <- predict(logistic_train, newdata = train_data, type = "response")
  predicted_classes_train <- ifelse(logistic_predict_train > threshold, "Snow", "Deficit")
  
  logistic_predict_test <- predict(logistic_train, newdata = test_data, type = "response")
  predicted_classes_test <- ifelse(logistic_predict_test > threshold, "Snow", "Deficit")
  
  cm_test <- confusionMatrix(as.factor(predicted_classes_test), as.factor(test_data$snow_prediction), 
                             positive = "Snow")
  
  cm_train <- confusionMatrix(as.factor(predicted_classes_train), as.factor(train_data$snow_prediction), 
                              positive = "Snow")
  
  return(list(cm_train = cm_train, cm_test = cm_test))
}

results <- logistic_model_evaluation(data_train, data_test, snow_prediction ~ .)

sensitivity_train <- results$cm_train$byClass["Sensitivity"]
specificity_train <- results$cm_train$byClass["Specificity"]
accuracy_train <- results$cm_train$overall["Accuracy"]

sensitivity_test <- results$cm_test$byClass["Sensitivity"]
specificity_test <- results$cm_test$byClass["Specificity"]
accuracy_test <- results$cm_test$overall["Accuracy"]

table_results <- data.frame(
  Metric = c("Sensitivity", "Specificity", "Accuracy"),
  Training_Set = c(sensitivity_train, specificity_train, accuracy_train),
  Test_Set = c(sensitivity_test, specificity_test, accuracy_test)
)

kable(table_results, caption = "Model Performance Metrics")
```

##Logistic regression on data after oversampling
```{r}

oversampled_data_train$snow_prediction <- factor(oversampled_data_train$snow_prediction, levels = c("Snow", "Deficit"))
predicted_classes_train_over <- factor(predicted_classes_train_over, levels = c("Snow", "Deficit"))

cm <- confusionMatrix(predicted_classes_train_over, oversampled_data_train$snow_prediction, 
                      positive = "Snow")

print(cm)
```
```{r}
logistic_predict_test_over <- predict(logistic_train_over, newdata = data_test, type = "response")

predicted_classes_test_over <- ifelse(logistic_predict_test_over > 0.5, "Snow", "Deficit")

data_test$snow_prediction <- factor(data_test$snow_prediction, levels = c("Snow", "Deficit"))
predicted_classes_test_over <- factor(predicted_classes_test_over, levels = c("Snow", "Deficit"))

cm_test_over <- confusionMatrix(as.factor(predicted_classes_test_over), as.factor(data_test$snow_prediction), 
                                positive = "Snow")
print(cm_test_over)





```

```{r, echo=FALSE}
tree_train_gini <- rpart(snow_prediction ~ ., data = oversampled_data_train, method = "class", parms = list(split = "gini"), maxdepth = 4)
```

```{r, echo=FALSE}
tree_predict_test_gini <- predict(tree_train_gini, newdata = data_test, type = "class")

confusionMatrix(tree_predict_test_gini, data_test$snow_prediction, positive ="Snow")
```

```{r, echo=FALSE}
rpart.plot(tree_train_gini)
```

```{r}
tree_gini <- rpart(snow_prediction ~ ., data = oversampled_data_train, method = "class", parms = list(split = "gini"), maxdepth = 4)

tree_predict_test_gini <- predict(tree_train_gini, newdata = data_test, type = "class")

conf_matrix <- confusionMatrix(tree_predict_test_gini, data_test$snow_prediction, positive = "Snow")

accuracy <- conf_matrix$overall['Accuracy']
specificity <- conf_matrix$byClass['Specificity']
sensitivity <- conf_matrix$byClass['Sensitivity']

cat("Accuracy: ", round(accuracy, 3), "\n")
cat("Specificity: ", round(specificity, 3), "\n")
cat("Sensitivity: ", round(sensitivity, 3), "\n")

```


### SVM Model

A detailed analysis of three support vector machine models with different kernels: linear, radial and polynomial was performed. All models were trained to predict the occurrence of snowfall using the regularization cost parameter set to 10. The performance was thoroughly assessed and presented using a confusion matrixes, allowing for a deeper understanding of their performance in various aspects of classification. <br>
The SVM model with a polynomial kernel has the highest accuracy, specificity and sensitivity. That suggests the highest efficiency in identifying snow days and makes it the preferred model for the given problem.

```{r, echo=FALSE}
svm_model_linear <- svm(snow_prediction ~ ., data = oversampled_data_train, kernel = "linear", cost = 10)

svm_model_radial <- svm(snow_prediction ~ ., data = oversampled_data_train, kernel = "radial", cost = 10, positive ="Snow")

svm_model_polynomial <- svm(snow_prediction ~ ., data = oversampled_data_train, kernel = "polynomial", cost = 10, positive ="Snow")
```

```{r, echo=FALSE}
svm_models <- list(svm_model_linear, svm_model_radial, svm_model_polynomial)
model_names <- c("Linear", "Radial", "Polynomial")

confusion_matrices <- list()

for (i in seq_along(svm_models)) {
  svm_predict_test <- predict(svm_models[[i]], newdata = data_test)
  
  confusion_matrices[[i]] <- confusionMatrix(svm_predict_test, data_test$snow_prediction, positive = "Snow")
}

for (i in seq_along(confusion_matrices)) {
  cat("Confusion Matrix for", model_names[i], "SVM Model:\n")
  print(confusion_matrices[[i]])
}
```

